An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2022-05-10 simplified_sdc_intro/paper.tex
more edits

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute during explosive burning or at high
temperatures where nuclear statistical equilibrium is established.  We
develop a simplified approach based on the ideas of spectral deferred
corrections (SDC) coupling of explicit hydrodynamics and stiff
reaction sources as an alternative to operator splitting or the more
comprehensive SDC approach we demonstrated previously.  We apply the
new method to some example problems with small and moderately-sized
astrophysical nuclear reaction networks and explore the timestep size
and reaction network tolerances to show that the simplified-SDC
approach compares favorably with traditional Strang operator
splitting.  This is all done in the framework of the
\castro\ hydrodynamics code, and all algorithm implementations are
freely available.
\end{abstract}

@@

In Strang splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and showed that integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  A variation on Strang splitting called (re-)balanced
splitting was developed in \citet{speth:2013}, but we do not explore
that here. \MarginPar{keep this sentence?}

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach (see \citealt{ppmunsplit}).  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).
Usually these limiters are used in a reactive manner---the next step will see
a smaller timestep based on what happened during the current step, but the
current step is still accepted, even if it violated these conditions.

\castro\ has the ability to reject a timestep if it detects a failure and retry
with smaller timesteps (subcycling to make up the original required timestep).
Among the conditions that can trigger this are density falling below zero during
advection, the ODE integration failing to converge in the implicit solve (due to
too many steps or the internal timestep falling below a minimum), or violation
of one of the timestep limiters during the step.  This means that the timestep
constraints are proactive instead of reactive.  The retry mechanism in \castro\
works with both the Strang and simplified-SDC integration scheme.  Retrying the
step is much more stringent that simply adjusting the next step, so we put a cap
on the number of retries allowed, aborting if we exceed the cap.

@@

The basic idea of spectral deferred corrections \MarginPar{ref} is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.  This is the
classic approach that we implemented in \cite{castro_sdc}.

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar (but unpublished) implementation in the \maestroex\ simulation code \cite{maestroex}.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
In this approach, we
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.   The overall flow of the simplified-SDC algorithm is shown
in Figure~\ref{fig:flowchart_II}.

@@

For Strang, we require that
an individual mass fraction not change too much over a step:
\begin{equation}
\label{eq:vode_species_reject}
\begin{array}{c}|X_k^{m+1}| < f |X_k^m| \\ |X_k^{m+1}| > \frac{1}{f} |X_k^m| \end{array} ~  \mbox{if}~ |X_k^m| > \eta \atolspec ~\mbox{and}~ |X_k^{m+1}| > \eta \atolspec
\end{equation}
Here, $m$ is the current VODE solution and $m+1$ is the potential
new-time solution.  The parameter $f$ is the factor a mass fraction is
allowed to change per VODE step.  We use $f = 4$ for the results here.
The parameter $\eta$ allows us to only use this condition for species
with mass fractions above a threshold.  We use $\eta = 1$ by default,
but $\eta = f$ sometimes helps prevent VODE from worrying about
species that dip below the threshold during integration, since VODE
uses a norm over all of the integrated variables to compute the total
error.  If these conditions are violated, then VODE will reject the
timestep and retry with a smaller step.  We also require that the mass
fractions are contained in $[0, 1]$ to a tolerance of $10^{-2}$.
For SDC, we enforce these same constraints, but for
Eq.~\ref{eq:vode_species_reject}, we only use the change from $m$ to $m+1$ due
to reactions, by subtracting off the advective contribution over that substep.
We also require that:
\begin{align}
\rho^{m+1} &> 0 \\
(\rho e)^{m+1} &> 0
\end{align}

Finally, some of the rates in \citet{reaclib} increase by 100s of orders of
magnitude at low temperature, presumably because the interpolant fit was done at
higher temperatures.  Since VODE can take exploratory righthand side evaluations
outside of the nominal time range, it can encounter these poorly-behaved rates
and it can cause the integration to fail (by generating {\tt inf}s).  To prevent
this behavior, we set a lower temperature limit of the reaction rates, below
which we zero them out.  For the simulations here, we choose a cutoff of
$5\times 10^7~\mathrm{K}$.

@@

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  This is based on the acoustic
pulse problem from \citep{mccorquodalecolella}, and was also used in
\citep{strang_rnaas}.  A periodic domain with a uniform entropy is
initialized with a pressure profile with a small perturbation.  This
drives a low amplitude acoustic wave radially outward from the center.
A simple reaction network with the triple-alpha and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ rates releases energy.  By
running at different resolutions, with a fixed timestep that scales
with resolution, we can compute the convergence rate.  In order to
demonstrate second-order convergence, the problem must be smooth (so
the slope limiters don't kick in severely), which means that the
energy release cannot be so vigorous as to drive shocks.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang and simplified-SDC
convergence respectively.  We see that both methods converge nearly
second order, and agree well with each other.  This demonstrates that both
methods are working as designed.

\subsection{Sub-Chandra Double Detonation}

@@

We do a suite of runs with both Strang and the simplified-SDC integration, using
CFL numbers of 0.1, 0.2, and 0.4 and reaction network tolerances of $\atolspec =
\rtolspec = 10^{-5}$ and $\atolener = \rtolener = 10^{-5}$.  We also run the CFL
= 0.2 case with a tighter tolerance, using  $\atolspec = 10^{-8}$.  For the step
rejection part of VODE, we use $f = 4$ and $\eta = 4$.

Figure \ref{fig:subch_strang_cfl02} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the strang
CFL = 0.82 simulation ({\tt strang\_subch2\_cfl0.2}).  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.

\begin{figure}
\centering
\plotone{strang_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_strang_cfl02}}
\end{figure}

We next look at the simplified-SDC version of this same case, {\tt
  sdc\_subch2\_cfl0.2}.  Figure~\ref{fig:subch_sdc_cfl02} shows the
temperature, mean molecular weight, and energy generation rate.
Compared to Figure~\ref{fig:subch_strang_cfl02}, we see that the
composition and energy generation rates are smooth, without any of the
checkerboarding that plagued the Strang simulation.
This suggests that the simplified-SDC algorithm has an easier time
attaining the correct equilibrium in these conditions.

\begin{figure}
\centering
\plotone{sdc_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_sdc_cfl02}}
\end{figure}

We now look at the evolution of the state while burning by storing the
thermodynamic state internal to VODE each substep as it evaluates the
righthand side of the ODE system it is integrating.  We pick a point
on the symmetry axis a few zones behind the inward moving shock and
ran both a Strang and simplified-SDC simulation for a single
coarse-grid timestep.  To ensure that we start off the same, both
simulations were started from a checkpoint file from the
simplified-SDC simulation a few steps before 0.1~s.
Figure~\ref{fig:nonaka_plot} shows the evolution of \isot{Ni}{56} over
a single coarse-grid timestep (4 fine-grid timesteps).  We see that
the Strang solution is much more discontinuous, as the reactions and
advection push the solution in different ways, because the processes
are completely independent of one another.  If an equilibrium state is
important to the nucleosynthesis, the large deviations in the Strang
solution mean that we are evaluating the reaction network with a state
out of equilibrium for much of the integration time. \MarginPar{need to double check the data---why are there points in the vertical part of Strang?}

\begin{figure}
\plotone{sdc_compare}
\caption{\label{fig:nonaka_plot} Evolution of $X(\isotm{Ni}{56})$ over
  4 fine-grid timesteps in a single zone behind the inward propagating
  shock.  The Strang points (blue) show the first $\Delta t/2$ burn
  bring the abundance away from equilibrium, then the advection
  overcorrecting it, and finally the second $\Delta t/2$ bringing it
  back toward the equilibrium.  The two simplified-SDC iterations
  appear much smoother.  For the SDC runs, there are a few points
  evaluated beyond the stop time---VODE can use data from outside the
  integration range to interpolate to the final solution.}
\end{figure}

We next look at how the structure seen in
Figures~\ref{fig:subch_strang_cfl02} and \ref{fig:subch_sdc_cfl02}
vary with CFL number and network tolerance.  We focus just on the
region behind the inward propagating shock.
Figures~\ref{fig:strang_enuc_summary} and
\ref{fig:strang_abar_summary} show the nuclear energy generation rate
and mean molecular weight for all of the Strang runs.  We see that
changing the CFL number does little to affect the state behind the
shock---it continues to take on the mottled appearance characteristic
of not reaching a proper equilibrium.  For the case with tight tolerances, we
see that the state is greatly improved.

@@

The SDC comparisons are shown in Figures~\ref{fig:sdc_enuc_summary}
and \ref{fig:sdc_abar_summary}.  We see that the state behind the
shock looks great regardless of the CFL or integrator tolerances.

@@

We presented a simplified spectral deferred corrections scheme for coupling
hydrodynamics and reactions.  We demonstrated that for a moderate-sized reaction
network, the simplified-SDC method compares well against Strang for the double
detonation problem---getting a better solution at looser reaction tolerances,
and consistent results across a range of CFL numbers.  The benefits of this new
integration scheme are likely problem and network dependent, and we will explore
it on other science problemsin future papers.

We have not focused on performance or load-balancing the reactions in
the tests that we ran, but for the simulations shown here,
the simplified-SDC method runs as fast (or faster) than the Strang method
for the same tolerances and CFL, despite having to do the advection and reactions
twice.  We would expect similar behavior for any simulations that are
dominated by the cost of reactions.  If the results above a robust across
different problems and networks, then this suggests that we can run
the SDC simulation with larger CFL and/or less strict reaction tolerances,
which would translate into a large performance boost.

This new time-integration method is freely available in \castro, and
we will continue to explore it on other problems.  We are particularly
interested in massive star evolution with this method, where the
vigorous shell burning leading up to core-collapse would be an ideal
application for the simplified-SDC method.  In a follow-on paper, we
will show how to couple this new integration method to a table for
evaluating an NSE distribution in the iron core of a massive star.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute during explosive burning or at high
temperatures where nuclear statistical equilibrium is established.  We
develop a simplified approach based on the ideas of spectral deferred
corrections (SDC) coupling of explicit hydrodynamics and stiff
reaction sources as an alternative to operator splitting or the more
comprehensive SDC approach we demonstrated previously.  We apply the
new method to a double detonation problem with a moderately-sized
astrophysical nuclear reaction network and explore the timestep size
and reaction network tolerances to show that the simplified-SDC
approach compares favorably with traditional Strang operator
splitting.  This is all done in the framework of the
\castro\ hydrodynamics code, and all algorithm implementations are
freely available.
\end{abstract}

@@

In the Strang splitting~\cite{strang:1968} flavor of operator splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

@@

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  A variation on Strang splitting called (re-)balanced
splitting was developed in \citet{speth:2013}, but we do not explore
that here. \MarginPar{keep this sentence?}

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach (see \citealt{ppmunsplit}).  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).
Usually these limiters are used in a reactive manner---the next step will see
a smaller timestep based on what happened during the current step, but the
current step is still accepted, even if it violated these conditions.  Our
goal is to avoid the need for these limiters by doing the coupling of hydrodynamics
and reactions better.

\castro\ has the ability to reject a timestep if it detects a failure and retry
with smaller timesteps (subcycling to make up the original required timestep).
Among the conditions that can trigger this are density falling below zero during
advection, the ODE integration failing to converge in the implicit solve (due to
too many steps or the internal timestep falling below a minimum), or violation
of one of the timestep limiters during the step.  This means that the timestep
constraints are proactive instead of reactive.  The retry mechanism in \castro\
works with both the Strang and simplified-SDC integration scheme.  Retrying the
step is much more stringent that simply adjusting the next step, so we put a cap
on the number of retries allowed, aborting if we exceed the cap. \MarginPar{describe in more detail?}

@@

The basic idea of spectral deferred corrections \MarginPar{ref} is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.  This is the
classic approach that we implemented in Castro in \citet{castro_sdc}.

Our simplified-SDC approach is based on the method described in \citet{SDC-old}
and a similar (but unpublished) implementation in the \maestroex\ simulation
code \citep{maestroex}. We again start by considering the update in integral
form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
In this approach, we
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.   The overall flow of the simplified-SDC algorithm is shown
in Figure~\ref{sdc_flowchart}.

@@

For Strang, we require that
an individual mass fraction not change too much over a step:
\begin{equation}
\label{eq:vode_species_reject}
\begin{array}{c}|X_k^{m+1}| < f |X_k^m| \\ |X_k^{m+1}| > \frac{1}{f} |X_k^m| \end{array} ~  \mbox{if}~ |X_k^m| > \eta \atolspec ~\mbox{and}~ |X_k^{m+1}| > \eta \atolspec
\end{equation}
Here, $m$ is the current VODE solution and $m+1$ is the potential
new-time solution.  The parameter $f$ is the factor a mass fraction is
allowed to change per VODE step.  We use $f = 4$ for the results here.
The parameter $\eta$ allows us to only use this condition for species
with mass fractions above a threshold.  We use $\eta = 1$ by default,
but $\eta = f$ sometimes helps prevent VODE from worrying about
species that dip below the threshold during integration, since VODE
uses a norm over all of the integrated variables to compute the total
error.  If these conditions are violated, then VODE will reject the
timestep and retry with a smaller step.  We also require that the mass
fractions are contained in $[0, 1]$ to a tolerance of $10^{-2}$.
For SDC, we enforce these same constraints, but for
Eq.~\ref{eq:vode_species_reject}, we only use the change from $m$ to $m+1$ due
to reactions, by subtracting off the advective contribution over that substep.
Finally, we enforce that:
\begin{align}
\rho^{m+1} &> 0 \\
(\rho e)^{m+1} &> 0
\end{align}

Some of the rates in \citet{reaclib} increase by 100s of orders of
magnitude at low temperature, presumably because the interpolant fit was done at
higher temperatures.  Since VODE can take exploratory righthand side evaluations
outside of the nominal time range, it can encounter these poorly-behaved rates
and it can cause the integration to fail (by generating {\tt inf}s).  To prevent
this behavior, we set a lower temperature limit of the reaction rates, below
which we zero them out.  For the simulations in section~\ref{sec:dd}, we choose a cutoff of
$5\times 10^7~\mathrm{K}$.

@@

\citet{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  This is based on the acoustic
pulse problem from \citet{mccorquodalecolella}, and was also used in
\citet{strang_rnaas}.  A periodic domain with a uniform entropy is
initialized with a pressure profile with a small perturbation.  This
drives a low amplitude acoustic wave radially outward from the center.
A simple reaction network with the triple-alpha and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ rates releases energy.  By
running at different resolutions, with a fixed timestep that scales
with resolution, we can compute the convergence rate.  In order to
demonstrate second-order convergence, the problem must be smooth (so
the slope limiters don't kick in severely), which means that the
energy release cannot be so vigorous as to drive shocks.

@@

Tables \ref{table:react_converge_strang} and \ref{table:react_converge_sdc} show
the Strang and simplified-SDC convergence respectively.  The convergence rate is
is computed by taking log-base-2 of the ratio of adjacent errors.  A rate of 2.0
indicates second order convergence.  We see that both methods converge nearly
second order, and agree well with each other.  This demonstrates that both
methods are working as designed.

\subsection{Sub-Chandra Double Detonation}
\label{sec:dd}

@@

We do a suite of runs with both Strang and the simplified-SDC integration, using
CFL numbers of 0.1, 0.2, and 0.4 and reaction network tolerances of $\atolspec =
\rtolspec = 10^{-5}$ and $\atolener = \rtolener = 10^{-5}$.  We also run the CFL
= 0.2 case with a tighter tolerance, using  $\atolspec = 10^{-8}$.  For the step
rejection part of VODE, we use $f = 4$ and $\eta = 4$.  The runs are labeled with
the prefix {\tt sdc\_} or {\tt strang\_} to detonate the integrator, and also include
the CFL number in the name.  Finally, the runs with the tighter tolerances have
the suffix {\tt \_tol1.e-8}.  We note that both the Strang and the SDC runs encounter
VODE integration failures occassionally and utilize the \castro\ step retry functionality.

Figure \ref{fig:subch_strang_cfl02} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the strang
CFL = 0.2 simulation ({\tt strang\_subch2\_cfl0.2}).  At this point
in the evolution, the helium detonation has wrapped more than halfway
across the surface and the ingoing carbon detonation is approaching the
center of the star.  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.

\begin{figure}
\centering
\plotone{strang_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_strang_cfl02} Temperature, mean molecular weight, and
energy generation rate for the Strang CFL = 0.2 run with normal tolerances ({\tt
strang_subch2\_cfl0.2}) at 0.1 s.}
\end{figure}

We next look at the simplified-SDC version of this same case, {\tt
sdc\_subch2\_cfl0.2}.  Figure~\ref{fig:subch_sdc_cfl02} shows the temperature,
mean molecular weight, and energy generation rate. Compared to
Figure~\ref{fig:subch_strang_cfl02}, we see that the composition and energy
generation rates are smooth, without any of the checkerboarding that plagued the
Strang simulation.  This suggests that the simplified-SDC algorithm has an
easier time attaining the correct equilibrium in these conditions. More
strikingly, we see that the inward propagating carbon detonation has not
progressed as far into the white dwarf at this time.  The two integration
methods are giving very different results here.

\begin{figure}
\centering
\plotone{sdc_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_sdc_cfl02} Temperature, mean molecular weight, and
energy generation rate for the simplified-SDC CFL = 0.2 run with normal tolerances ({\tt
sdc_subch2\_cfl0.2}) at 0.1 s.}
\end{figure}

% We now look at the evolution of the state while burning by storing the
% thermodynamic state internal to VODE each substep as it evaluates the
% righthand side of the ODE system it is integrating.  We pick a point
% on the symmetry axis a few zones behind the inward moving shock and
% ran both a Strang and simplified-SDC simulation for a single
% coarse-grid timestep.  To ensure that we start off the same, both
% simulations were started from a checkpoint file from the
% simplified-SDC simulation a few steps before 0.1~s.
% Figure~\ref{fig:nonaka_plot} shows the evolution of \isot{Ni}{56} over
% a single coarse-grid timestep (4 fine-grid timesteps).  We see that
% the Strang solution is much more discontinuous, as the reactions and
% advection push the solution in different ways, because the processes
% are completely independent of one another.  If an equilibrium state is
% important to the nucleosynthesis, the large deviations in the Strang
% solution mean that we are evaluating the reaction network with a state
% out of equilibrium for much of the integration time. \MarginPar{need to double check the data---why are there points in the vertical part of Strang?}

% \begin{figure}
% \plotone{sdc_compare}
% \caption{\label{fig:nonaka_plot} Evolution of $X(\isotm{Ni}{56})$ over
%   4 fine-grid timesteps in a single zone behind the inward propagating
%   shock.  The Strang points (blue) show the first $\Delta t/2$ burn
%   bring the abundance away from equilibrium, then the advection
%   overcorrecting it, and finally the second $\Delta t/2$ bringing it
%   back toward the equilibrium.  The two simplified-SDC iterations
%   appear much smoother.  For the SDC runs, there are a few points
%   evaluated beyond the stop time---VODE can use data from outside the
%   integration range to interpolate to the final solution.}
% \end{figure}

We next look at how the structure seen in
Figures~\ref{fig:subch_strang_cfl02} and \ref{fig:subch_sdc_cfl02}
vary with CFL number and network tolerance.  We focus just on the
region behind the inward propagating shock.
Figures~\ref{fig:strang_enuc_summary} and
\ref{fig:strang_abar_summary} show the nuclear energy generation rate
and mean molecular weight for all of the Strang runs.  We see that
changing the CFL number does little to affect the state behind the
shock---it continues to take on the mottled appearance characteristic
of not reaching a proper equilibrium.  For the case with tight tolerances, we
see that the state is greatly improved.  Furthermore, it gets the inward
propagating carbon detonation in the same position as the SDC runs.

@@

The SDC comparisons are shown in Figures~\ref{fig:sdc_enuc_summary}
and \ref{fig:sdc_abar_summary}.  We see that the state behind the
shock looks great regardless of the CFL or integrator tolerances.
This demonstrates that the simplified-SDC integration algorithm is more
robust than Strang integration for this problem.

@@

We presented a simplified spectral deferred corrections scheme for coupling
hydrodynamics and reactions.  We demonstrated that for a moderate-sized reaction
network, the simplified-SDC method compares well against Strang for the double
detonation problem---getting a better solution at looser reaction tolerances,
and consistent results across a range of CFL numbers.  The benefits of this new
integration scheme are likely problem and network dependent, and we will explore
it on other science problems in future papers.

We have not focused on performance or load-balancing the reactions in
the tests that we ran, but for the simulations shown here,
the simplified-SDC method runs as fast (or faster) than the Strang method
for the same tolerances and CFL number, despite having to do the advection and reactions
twice.  We would expect similar behavior for any simulations that are
dominated by the cost of reactions.  If the results above are robust across
different problems and networks, then this suggests that we can run
the SDC simulation with larger CFL and/or less strict reaction tolerances,
which would translate into a large performance boost.

This new time-integration method is freely available in \castro, and
we will continue to explore it on other problems.  Although these simulations
were run on CPUs, the entire simplified-SDC algorithm also runs on GPUs.
interested in massive star evolution with this method, where the
vigorous shell burning leading up to core-collapse would be an ideal
application for the simplified-SDC method.  In a follow-on paper, we
will show how to couple this new integration method to a table for
evaluating an NSE distribution in the iron core of a massive star.
